\textbf{A framework for effective application of machine learning to microbiome-based classification problems}

### 2020-01-28 submission/Table_S1.tex

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={A framework for effective application of machine learning to microbiome-based classification problems},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\title{\textbf{A framework for effective application of machine learning to
microbiome-based classification problems}}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\vspace{35mm}

Running title: Machine learning framework to model microbiome data

\vspace{35mm}

Begüm D. Topçuoğlu\({^1}\), Nicholas A. Lesniak\({^1}\), Mack
Ruffin\({^3}\), Jenna Wiens\({^2}\), Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

\vspace{38mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}},
\href{mailto:wiensj@umich.edu}{\nolinkurl{wiensj@umich.edu}}

1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI 48109

2. Department of Electrical Engineering and Computer Science, University
of Michigan, Ann Arbor, MI 48109

3. Department of Family Medicine and Community Medicine, Penn State
Hershey Medical Center, Hershey, PA

@@

\linenumbers

\subsection{Abstract}\label{abstract}

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in the diagnosis of many
diseases such as inflammatory bowel disease, diabetes, and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate, and
interpret ML models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immediately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and was
inherently interpretable. Our analysis highlights the importance of
choosing an ML approach based on the goal of the study, as the choice
will inform expectations of performance and interpretability.

\subsection{Importance}\label{importance}

Prediction of diseases using machine learning (ML) is rapidly being
adopted in microbiome studies. However, the estimated performance
associated with these ML models is likely over-optimistic. Moreover,
there is a trend towards using black box models without a discussion of
the difficulty of interpreting such models when trying to identify
microbial biomarkers of disease. This work represents a step towards
developing more reproducible ML practices in applying ML to microbiome
research. We implement a rigorous pipeline and emphasize the importance
of selecting ML models that reflect the goal of the study. These
concepts are not particular to the study of human health but can also be
applied to environmental microbiology studies.

\subsection{Background}\label{background}

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
disease. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. For example it is not possible to
classify individuals as having healthy colons or colorectal lesions
using Bray-Curtis distances based on the 16S rRNA gene sequences
collected from fecal samples {[}Figure S1{]}. This variation is likely
due to the ability of many bacterial populations to fill the same niche
such that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population separately (1). Recently,
machine learning (ML) models have grown in popularity among microbiome
researchers as our ability to sample large numbers of individuals has
grown; such models can effectively account for the interpersonal
microbiome variation and the ecology of disease.

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without separate held-out test data; unreported variation between
the predictive performance on different folds of cross-validation; and
unreported variation between cross-validation and testing performances.
Though the microbiome field is making progress to avoid some of these
pitfalls including validating their models on independent datasets (8,
19, 20) and introducing accessible and open-source ML tools (21--24),
more work is needed to improve reproducibility further and minimize
overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modeling approach has often been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
non-linear models such as random forest and deep neural networks (3, 12,
25--27) over simpler models such as logistic regression or other linear
models (19, 23, 28). Although in some cases, complex models may capture
important non-linear relationships and therefore yield better
predictions, they can also result in black boxes that lack
interpretability. Such models require post hoc explanations to quantify
the importance of each feature in making predictions. Depending on the
goal of the modeling, other approaches may be more appropriate. For
example, researchers trying to identify the microbiota associated with
disease may desire a more interpretable model, whereas clinicians may
emphasize predictive performance. Nonetheless, it is essential to
understand that the benefit of more complex, less interpretable models
may be minimal (29--31). It is important for researchers to justify
their choice of modeling approach.

In this study, we provided steps toward standardization of machine
learning methods for microbiome studies which are often poorly
documented and executed. To showcase a rigorous ML pipeline and to shed
light on how ML model selection can affect modeling results, we
performed an empirical analysis comparing the predictive performance,
interpretability and training times of seven modeling approaches with
the same dataset and pipeline. We built three linear models with
different forms of regularization: L2-regularized logistic regression
and L1 and L2-regularized support vector machines (SVM) with a linear
kernel. We also trained four non-linear models: SVM with radial basis
function kernel, a decision tree, random forest, and gradient boosted
trees. We compared their predictive performance, interpretability, and
training time. To demonstrate the performance of these modeling
approaches and our pipeline on a case study, we used a data from a
previously published study that sought to classify individuals as having
healthy colons or colonic lesions based on the 16S rRNA gene sequences
collected from fecal samples (4). This dataset was selected because it
is a relatively large collection of individuals (N=490) connected to a
clinically significant disease where there is ample evidence that the
disease is driven by variation in the microbiome (2, 4, 5, 32). With
this dataset, we developed an ML pipeline that can be used in many
different scenarios for training and evaluating models. This framework
can be easily applied to other host-associated and environmental
microbiome datasets. We also provided an aspirational rubric for
evaluating the rigor of ML practices applied to microbiome data {[}Table
S1{]} to urge the audience to be diligent with with their study design,
model selection, development, evaluation and interpretation.

\subsection{Results}\label{results}

\textbf{Model selection and pipeline construction}. We established a
reusable ML pipeline for model selection and evaluation, focusing on
seven different commonly used supervised learning algorithms {[}Figure
1, Table 1{]}.

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset, while the test
set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
distribution of controls and cases found in the full dataset, we
performed stratified splits. For example, our full dataset included 490
individuals. Of these, 261 had healthy colons (53\%) and 229 had a
screen relevant neoplasia (SRN; 46.7\%). A training set included 393
individuals, of which 209 had an SRN (53\%), while the test set was
composed of 97 individuals, of which 52 had an SRN (54\%). The training
data were used to build and select the models, and the test set was used
for evaluating the model. We trained seven different models using the
training data {[}Table 1{]}.

Model selection requires tuning hyperparameters. Hyperparameters are
parameters that need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S2 and S3{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that the model's predictions are no
different than random. To select hyperparameters, we performed a grid
search for hyperparameter settings when training the models. Default
hyperparameter settings in developed ML packages available in R, Python,
and MATLAB programming languages may be inadequate for effective
application of classification algorithms and need to be optimized for
each new ML task. For example, L1-regularized SVM with linear kernel
showed large variability between different regularization strengths (C)
and benefited from tuning as the default C parameter was 1 {[}Figure
S2{]}.

Once hyperparameters were selected, we trained the model using the full
training dataset and applied the final model to the held-out data to
evaluate the testing predictive performance of each model. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to obtain a robust interpretation of model
performance, less likely to be affected by a ``lucky'' or ``unlucky''
split {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
The predictive performance of random forest model was higher than other
ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5; The p-value was manually calculated using the
sampling distribution of the test statistic under the null hypothesis)
(Figure S4). Similarly, L2-regularized logistic regression, XGBoost,
L2-regularized SVM with linear and radial basis function kernel AUROC
values were not significantly different from one another and had median
AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
results demonstrate that the most complex model (XGBoost) did not have
the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well as non-linear models.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
could indicate that the models were overfit to the training data. The
largest difference in median AUROCs was 0.021 in L1-regularized SVM with
linear kernel, followed by SVM with radial basis function kernel and
decision tree with a difference of 0.007 and 0.006, respectively
{[}Figure 2{]}. These differences were relatively small and gave us
confidence in our estimate of the generalization performance of the
models.

To evaluate the variation in the estimated performance, we calculated
the range of AUROC values for each model using 100 data-splits. The
range among the testing AUROC values within each model varied by 0.230
on average across the seven models. If we had only done a single split,
then there is a risk that we could have gotten lucky or unlucky in
estimating model performance. For instance, the lowest AUROC value of
the random forest model was 0.593, whereas the highest was 0.810. These
results showed that depending on the data-split, the testing performance
can vary {[}Figure 2{]}. Therefore, it is important to employ multiple
data splits when estimating generalization performance.

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S5{]}. The
variation in cross-validation performance within both models at smaller
sample sizes was larger than when the full collection of samples was
used to train and validate the models. Because of the high
dimensionality of the microbiome data (6920 OTUs), large sample sizes
can lead to better models.

\textbf{Interpretation of each ML model.} We often use ML models not
just to predict a health outcome, but also to identify potential
biomarkers for disease. Therefore, model interpretation becomes crucial
for microbiome studies. Interpretability is related to the degree to
which humans can understand the reasons behind a model prediction
(33--35). ML models often decrease in interpretability as they increase
in complexity. In this study, we used two methods to help interpret our
models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest-ranked OTUs were shared across these three models (e.g., OTUs
50, 426, 609, 822, 1239). The great benefit of this approach was knowing
the sign and magnitude of each OTU coefficient in the trained model.
This allowed us to immidiately interpret negative and positive
coefficient signs as protective and risk factors, respectively and the
magnitude as the impact of these factors. However, this approach is
limited to linear models or models with prespecified interaction terms.

Second, to analyze non-linear models, we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data
{[}Figure S6{]}. We then calculated how much the predictive performance
of the model (i.e., testing AUROC values) decreased when each OTU or
group of OTUs was randomly permuted. We ranked the OTUs based on how
much the median testing AUROC decreased when it was permuted; the OTU
with the largest decrease ranked highest {[}Figure 4{]}. Among the
twenty OTUs with the largest impact, there was only one OTU (OTU 822)
that was shared among all of the models; however, we found three OTUs
(OTUs 58, 110, 367) that were important in each of the tree-based
models. Similarly, the random forest and XGBoost models shared four of
the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis
results also revealed that with the exception of the decision tree
model, removal of any individual OTU had minimal impact on model
performance. For example, if OTU 367 was permuted across the samples in
the decision tree model, the median AUROC dropped from 0.601 to 0.525.
In contrast, if the same OTU was permuted in the random forest model,
the AUROC only dropped from 0.695 to 0.680, which indicated high degree
of collinearity in the dataset. Permutation analysis allowed us to gauge
the importance of each OTU in non-linear models and partially account
for collinearity by grouping correlated OTUs to determine their impact
as a group.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S7{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S7{]}, 17 of the 20 top OTUs (e.g., OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. These results indicate that both methods are
consistent in selecting the most important OTUs.

We also compared the top 20 OTUs selected by permutation importance in
L2-regularized logistic regression and the highest performing tree-based
models, random forest and XGBoost. 2 and 5 OTUs, respectively, were
shared among the models. These results indicate that we were able to
identify important OTUs that are shared across the highest performing
linear and non-linear models when we use permutation importance as our
interpretation method.

We then evaluated the difference in relative abundances of the top 20
OTUs identified in L2-regularized logistic regression and random forest
models between healthy patients and patients with SRNs {[}Figure S8{]}.
There were minimal differences in relative abundances across OTUs
between different diagnoses. This supports our claim that it is not
possible to differentiate disease vs healthy states by focusing on
individual taxa.

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. The training times increased with
the complexity of the model and the number of potential hyperparameter
combinations. Also, the linear models trained faster than non-linear
models {[}Figures S1-S2; Figure 5{]}.

\subsection{Discussion}\label{discussion}

There is a growing awareness that many human diseases and environmental
processes are not driven by a single organism but are the product of
multiple bacterial populations. Traditional statistical approaches are
useful for identifying those cases where a single organism is associated
with a process. In contrast, ML methods offer the ability to incorporate
the structure of the microbial communities as a whole and identify
associations between community structure and disease state. If it is
possible to classify communities reliably, then ML methods also offer
the ability to identify those microbial populations within the
communities that are responsible for the classification. However, the
application of ML in microbiome studies is still in its infancy, and the
field needs to develop a better understanding of different ML methods,
their strengths and weaknesses, and how to implement them.

To address these needs, we developed an open-sourced framework for ML
models. Using this pipeline, we benchmarked seven ML models and showed
that the tradeoff between model complexity and performance may be less
severe than originally hypothesized. In terms of predictive performance,
the random forest model had the best AUROC compared to the other six
models. However, the second-best model was L2-regularized logistic
regression with a median AUROC difference of less than 0.015 compared to
random forest. While our implementation of random forest took 83.2 hours
to train, our L2-regularized logistic regression trained in 12 minutes.
In terms of interpretability, random forest is a non-linear ML model,
while L2-regularized logistic regression, a linear model, is easily
interpreted according to the feature weights. Comparing many different
models showed us that the most complex model was not necessarily the
best model for our ML task.

We established a pipeline that can be generalized to any modeling method
that predicts a binary health outcome. We performed a random data-split
to create a training set (80\% of the data) and a held-out test set
(20\% of the data), which we used to evaluate predictive performance. We
used the AUROC metric to evaluate predictive performance as it is a
clinically relevant evaluation metric for our study. We repeated this
data-split 100 times to measure the possible variation in predictive
performance. During training, we tuned the model hyperparameters with a
repeated five-fold cross-validation. Despite the high number of features
microbiome datasets typically have, the models we built with this
pipeline generalized to the held-out test sets.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures that create collinearity in the datasets. This can hinder our
ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (37). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other; however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. We are choosing to leave the decision to define
correlation and setting up correlations thresholds to the researchers
for their own microbiome questions. However, we would like to urge them
to investigate the correlation structures of the data to avoid
misinterpreting the models. This is likely to be a particular problem
with shotgun metagenomic datasets where collinearity will be more
pronounced due to many genes being correlated with one another because
they come from the same chromosome. True causal mechanisms (e.g., role
of microbiome in colorectal cancer) cannot be explained solely by the
highest performing machine learning model (38). To identify the true
underlying microbial factors of a disease, it is crucial to follow up on
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with healthy
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with healthy colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (39). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 40, 41), but microbiome datasets
often suffer from having many features and small sample sizes, which
result in overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct undertaking
(42). For example, we currently lack standardized methods to collect
patient samples, generate sequence data, and report clinical data. We
are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (43). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

Our study highlights the need to make educated choices at every step of
developing an ML model with microbiome data. We created an aspirational
rubric that researchers can use to identify potential pitfalls when
using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
have highlighted the trade-offs between model complexity and
interpretability, the need for tuning hyperparameters, the utility of
held-out test sets for evaluating predictive performance, and the
importance of considering correlation structures in datasets for
reliable interpretation. Furthermore, we underscored the importance of
proper experimental design and methods to help us achieve the level of
validity and accountability we want from models built for patient
health.

\subsection{Materials and Methods}\label{materials-and-methods}

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (44). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (44). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (45), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

\textbf{Model training and evaluation.} Models were trained using the
caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
calculate decision values for models generated using L2-regularized SVM
with linear kernel and L1-regularized SVM with linear kernel. The code
for these changes on L2-regularized SVM with linear kernel and
L1-regularized SVM with linear kernel models are available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
and at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
respectively.

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter.For
L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (48).

The computational burden during model training due to model complexity
was reduced by parallelizing segments of the ML pipeline. We
parallelized the training of each data-split. This allowed the 100
data-splits to be processed through the ML pipeline simultaneously at
the same time for each model. It is possible to further parallelize the
cross-validation step for each hyperparameter setting which we have not
performed in this study.

\textbf{Permutation importance workflow.} We calculated a Spearman's
rank-order correlation matrix and defined correlated OTUs as having
perfect correlation (correlation coefficient = 1 and p \textless{}
0.01). OTUs without a perfect correlation to each other were permuted
individually, whereas correlated ones were grouped together and permuted
at the same time.

\textbf{Statistical analysis workflow.} Data summaries, statistical
analysis, and data visualizations were performed using R (v.3.5.0) with
the tidyverse package (v.1.2.1). We compared the performance of the
models pairwise by calculating the difference between AUROC values from
the same data-split (for 100 data-splits). We determined if the models
were significantly different by calculating the empirical p-value (2 x
min(\% of AUROC differences \(\geq\) 0, \% of AUROC differences \(\leq\)
0) for the double tail event (e.g., Figure S4).

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at \url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/}.

\textbf{Acknowledgements.} We thank all the study participants of Great
Lakes-New England Early Detection Research Network. We would like to
thank the members of the Schloss lab for their valuable feedback. Salary
support for MR came from NIH grant 1R01CA215574. Salary support for PDS
came from NIH grants P30DK034933 and 1R01CA215574.

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-segata_metagenomic_2011}{}
1. \textbf{Segata N}, \textbf{Izard J}, \textbf{Waldron L},
\textbf{Gevers D}, \textbf{Miropolsky L}, \textbf{Garrett WS},
\textbf{Huttenhower C}. 2011. Metagenomic biomarker discovery and
explanation. Genome Biol \textbf{12}:R60.
doi:\href{https://doi.org/10.1186/gb-2011-12-6-r60}{10.1186/gb-2011-12-6-r60}.

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\author{}
\date{\vspace{-2.5em}}

\begin{document}

@@

\captionsetup{labelformat=empty}
\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
\small

\begin{tabular}{|l|l|l|l|}
\hline

\rowcolor{lightgray}
\textbf{Practice} & \textbf{Poor} & \textbf{Good} & \textbf{Better} \\ \hline

\makecell[l]{Source \\ of data} & \makecell[l]{Data do not reflect intended \\  application (e.g., data pertain \\ to only patients with carcinomas \\ but model is expected to \\ predict advanced adenomas).} & \makecell[l]{Data are appropriate \\ for intended application.} & \makecell[l]{Data reflect intended \\ use and will persist \\ (e.g., same OTU assignments \\ for new fecal samples).} \\ \hline

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population to which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline

\makecell[l]{Model \\ selection} & \makecell[l]{No justification for \\ classification method.} & \makecell[l]{Model choice is justified \\ for intended application.} & \makecell[l]{Different modeling choices \\ (justified for intended \\ application) are tested.} \\ \hline

\makecell[l]{Model \\ development} & \makecell[l]{No hyperparameter tuning.} & \makecell[l]{Different hyperparameter \\ settings are explored \\ on training data.} & \makecell[l]{Hyperparameter grid search \\ performed by cross-validation \\ on the training set.} \\ \hline

\makecell[l]{Model \\ evaluation} & \makecell[l]{Performance reported on the \\data used to train the model.} & \makecell[l]{Performance reported on \\ held-out test data.} & \makecell[l]{Performance reported on \\ multiple held-out test sets.} \\ \hline

\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g., when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline

\makecell[l]{Model \\ interpretation} & \makecell[l]{No model interpretation.} & \makecell[l]{Follow-up analyses to \\ determine what is driving \\ model performance.} & \makecell[l]{Hypotheses based on \\ feature importances \\ are generated and tested.} \\ \hline
